Data Specialization for a Staged Lambda Calculus

### 2015-07-06 paper/semantics.tex
a few more edits to the semantics

All three worlds contain unit, product, and function types defined in the usual fashion;
sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.
For example, given a term $e$ of type $A$ at world \bbtwo, 
$\next{e}$ has type $\fut A$ at \bbonem. 
This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
but it provides a handle (of type $\fut A$) to that eventual value which can be moved around now.
Computations at \bbonem\ can shuffle this handle around, but not access the future result it refers to.  
The only way to eliminate a $\fut$ wrapper is to do so from the future by using a {\tt prev}, yielding an $A$ at \bbtwo.

@@

@@

An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
projection, of type $\curr A$.

Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
open on $\curr{A}$ and $\curr{B}$ respectively.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$
and reduces all of its first-stage subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is a monostage object, this second pass is just standard monostage evaluation.
Consequently, this paper is concerned only with the first pass of evaluation.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

>>>>

All three worlds contain unit, product, and function types defined in the usual fashion;
sum and recursive types are provided for worlds \bbonep\ and \bbtwo\ 
(we omit them from \bbonem, because they are not needed for our examples).
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.

At the term level, {\tt next} blocks can be used to form future computations:
given a term $e$ of type $A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. 
This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
and it provides a handle (of type $\fut A$) now to that eventual value.
Computations at \bbonem\ can shuffle this handle around as a value, but the future result it refers to cannot be accessed.
This is because the only way to eliminate a $\fut$ wrapper is by using a {\tt prev}, which yields an $A$ at \bbtwo.
This feature was adapted from linear temporal logic, via \cite{davies96}.

@@

An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
projection, of type $\curr A$.
This elimination form, in contrast to that of $\fut$, does not permit world \bbonem\ subterms
within any world \bbonep\ term.

Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
open on $\curr{A}$ and $\curr{B}$ respectively.
Without this, there would be no way to get information out from under a $\curr$ wrapper.

@@

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is monostage by definition, this second pass is just standard unstaged evaluation.
Consequently, this paper does not describe in any detail the second pass of evaluation.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgement does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.
